Composing Compilers

### 2021-10-26 prisc.tex
Fix some typos

\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{stmaryrd}
\usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{cleveref}
\usepackage{listings}
\usepackage{mmmacros}
\usepackage[switch]{lineno}
%\modulolinenumbers[2]
\renewcommand{\linenumberfont}{\normalfont\bfseries\small\color{red}}

@@

\section{Compositionality}
Programs are written in some programming language.
In this work, programs are elements of $\partials$, which is the set of partial programs of a given programming language.
A compiler between languages $\S$ and $\T$ is a partial function $\stcomp{\bullet}$ from $\src{\partials}$ to $\trg{\partials}$.
To reason about preservation of program behavior (hyperproperties~\cite{clarkson08}), we have to define a suitable compiler property.

\begin{definition}[Robust Trace-Hyperproperty Preservation]\label{def:rtp}
  For a given class $\cC$, a compiler from languages $\S$ to $\T$ robustly preserves $\cC$ iff
  $$
  \forall\Pi\in\cC,\forall\src{p}\in\src{\partials},\rsat{\src{p}}{\Pi}\implies\rsat{\stcomp{\src{p}}}{\Pi}
  $$
  We write $\rtp{\stcomp{\bullet}}{\cC}$.
\end{definition}
\noindent
So, given any restriction on a program's behavior $\Pi$ and a source program $\src{p}$ that robustly satisfies $\Pi$ ($\rsat{\src{p}}{\Pi}$), the compiled program $\stcomp{\src{p}}$ also needs to robustly satisfy $\Pi$.
The intuition of \Cref{def:rtp} is as follows: If the programmer makes certain assumptions on what a program does, these assumptions should also hold for the compiled program.
% example (?)
Practical compilers are composed of multiple others.
Therefore, we now investigate compositionality of robust compilers.

@@

The following free theorem is a direct consequence of \Cref{lem:seqcompo} where the involved compilers' input and output are both partial programs in the same language.
Given that some set of compiler passes is robustly preserving properties, we can combine them in an arbitrary order and preserve the same least upper bound.
A compiler's pipeline ordering is difficult and often hand-tuned.
The lemma allows us to not care about the particular order of optimizations regarding their robust property preservation.
So, the compiler developer is free to swap passes around.
\begin{lemma}[Swappable]
  Given $\rtp{\ttcomp{\bullet}_{(1)}}{\cC_{1}}$ and $\rtp{\ttcomp{\bullet}_{(2)}}{\cC_{2}}$, then $\rtp{\ttcompN{\ttcomp{\bullet}_{(2)}}_{(1)}}{\cC_{1}\cap\cC_{2}}$ and $\rtp{\ttcompN{\ttcomp{\bullet}_{(1)}}_{(2)}}{\cC_{1}\cap\cC_{2}}$.
\end{lemma}

However in practice, compiler passes are not necessarily robustly preserving.
Consider the following C code snippet that performs an infinite loop if an invalid pointer is given:
\begin{lstlisting}[language=c]
int something(int* ptr) {
  while(!ptr);
  return *ptr;
}
\end{lstlisting}
Compiling such code with optimizations turned on yields an x86-program where the potentially infinite loop has been removed:
\begin{lstlisting}[language={[x86masm]Assembler}]
something(int*):
  mov eax, DWORD PTR [rdi]
  ret
\end{lstlisting}
Not only does the program terminate given an invalid pointer, it violates memory safety, which it did not before.

@@

We plan to extend our work with such source-code instrumentations which enforce classes of trace hyperproperties.
A sketch of the theory is as follows.

@@

\noindent
  We write $\sinstr{\stcomp{\bullet}}{\cC'}{\cC}$.
\end{definition}
We can now investigate an insecure compilation chain from e.g.\ memory-safe Rust to optimized, insecure C, to memory-safe CheckedC.
The intuition that a secure instrumentation saves the day does indeed hold and is captured by the following lemma.
\begin{lemma}[Final Enforcement]
  Given classes $\cC_{1},\cC_{2}$ and compilers $\sicomp{\bullet}$,$\itcomp{\bullet}$, we assume:

@@

Likewhise, running a compiler that does not respect e.g.\ memory-safety after a memory-safety instrumentation nullifies the preservation of memory-safety:
\begin{lemma}[Final Destination]
  Given classes $\cC_{1},\cC_{2}$ and compilers $\sicomp{\bullet}$,$\itcomp{\bullet}$, we assume:

@@

Now, consider any stereotypical compilation pipeline.
The security researcher has a strong interest to ensure that properties at source level are preserved at target level.
Thus, if source programs robustly satsify some property, so should the target.
However, it might not be necessary for compilation passes from one intermediate reperesentation to the other to preserve properties robustly, since compiler intermediate representations are outside the considered attack vector.
So, there might be some weaker property a pass should satisfy in order to render the whole compilation pipeline secure.

We intend to study property-free characterizations of the presented material.
Another area is the specialization of it to specific classes, such as $\cSafety$, and subsets thereof.

>>>>

\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{stmaryrd}
\usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{cleveref}
\usepackage{listings}
\usepackage{./../mmmacros}
\usepackage[switch]{lineno}
%\modulolinenumbers[2]
\renewcommand{\linenumberfont}{\normalfont\bfseries\small\color{red}}

@@

\section{Compositionality}
Programs are written in some programming language.
In this work, programs are elements of $\partials$, the set of partial programs of a given programming language.
A compiler between languages $\S$ and $\T$ is a partial function $\stcomp{\bullet}$ from $\src{\partials}$ to $\trg{\partials}$.
To reason about the preservation of program behaviour (hyperproperties~\cite{clarkson08}), we have to define a suitable compiler property.

\begin{definition}[Robust Trace-Hyperproperty Preservation]\label{def:rtp}
  For a given class $\cC$, a compiler from languages $\S$ to $\T$ robustly preserves $\cC$ iff
  $$
  \forall\Pi\in\cC,\forall\src{p}\in\src{\partials},\rsat{\src{p}}{\Pi}\implies\rsat{\stcomp{\src{p}}}{\Pi}
  $$
  We write $\rtp{\stcomp{\bullet}}{\cC}$.
\end{definition}
\noindent
So, given any restriction on a program's behavior $\Pi$ and a source program $\src{p}$ that robustly satisfies $\Pi$ ($\rsat{\src{p}}{\Pi}$), the compiled program $\stcomp{\src{p}}$ also needs to robustly satisfy $\Pi$.
The intuition of \Cref{def:rtp} is as follows: If the programmer makes certain assumptions on what a program does, these assumptions should also hold for the compiled program.
% example (?)
Practical compilers are composed of numerous others.
Therefore, we now investigate the compositionality of robust compilers.

@@

The following free theorem is a direct consequence of \Cref{lem:seqcompo} where the involved compilers' input and output are both partial programs in the same language.
Given that some compiler passes are robustly preserving properties, we can combine them in an arbitrary order and preserve the same least upper bound.
A compiler's pipeline ordering is difficult and often hand-tuned.
The lemma allows us to not care about the particular order of optimizations regarding their robust property preservation.
So, the compiler developer is free to swap passes around.
\begin{lemma}[Swappable]
  Given $\rtp{\ttcomp{\bullet}_{(1)}}{\cC_{1}}$ and $\rtp{\ttcomp{\bullet}_{(2)}}{\cC_{2}}$, then $\rtp{\ttcompN{\ttcomp{\bullet}_{(2)}}_{(1)}}{\cC_{1}\cap\cC_{2}}$ and $\rtp{\ttcompN{\ttcomp{\bullet}_{(1)}}_{(2)}}{\cC_{1}\cap\cC_{2}}$.
\end{lemma}

However, in practice, compiler passes are not necessarily robustly preserving.
Consider the following C code snippet that performs an infinite loop if an invalid pointer is given:
\begin{lstlisting}[language=c]
int something(int* ptr) {
  while(!ptr);
  return *ptr;
}
\end{lstlisting}
Compiling such code with optimizations turned on yields an x86-program where the potentially infinite loop has been removed:
\begin{lstlisting}[language={[x86masm]Assembler}]
something(int*):
  mov eax, DWORD PTR [rdi]
  ret
\end{lstlisting}
Not only does the program terminate given an invalid pointer, it violates memory safety, which it did not before.

@@

We plan to extend our work with such source-code instrumentations, which enforce classes of trace hyperproperties.
A sketch of the theory is as follows.

@@

\noindent
  We write $\sinstr{\stcomp{\bullet}}{\cC'}{\cC}$.
\end{definition}
We can now investigate an insecure compilation chain from e.g.\ memory-safe Rust to optimized, insecure C, to memory-safe CheckedC.
The intuition that a secure instrumentation saves the day holds and is captured by the following lemma.
\begin{lemma}[Final Enforcement]
  Given classes $\cC_{1},\cC_{2}$ and compilers $\sicomp{\bullet}$,$\itcomp{\bullet}$, we assume:

@@

Likewise, running a compiler that does not respect e.g.\ memory-safety after a memory-safety instrumentation nullifies the preservation of memory-safety:
\begin{lemma}[Final Destination]
  Given classes $\cC_{1},\cC_{2}$ and compilers $\sicomp{\bullet}$,$\itcomp{\bullet}$, we assume:

@@

Now, consider any stereotypical compilation pipeline.
The security researcher has a strong interest to ensure that properties at the source level are preserved at the target level.
Thus, if source programs robustly satisfy some property, so should the target.
However, it might not be necessary for compilation passes from one intermediate representation to the other to preserve properties robustly since compiler intermediate representations are outside the considered attack vector.
So, there might be some weaker property a pass has to satisfy in order to render the whole compilation pipeline secure.

We intend to study property-free characterizations of the presented material.
Another area is its specialization to specific classes, such as $\cSafety$, and subsets thereof.
